Analysis of competitive surfing tournaments with generalized Bradley-Terry

### 2021-08-21 inst/surfing_jsa.tex
more work

\title{Analysis of professional surfing tournaments with generalized Bradley-Terry}

\section*{Abstract}

@@

Paired comparisons are particularly favourable to Bradley-Terry
because the likelihood function is exponential family and as such is
amenable to analysis via general linear models~\cite{turner2012}.
However, this breaks down when more than two entities are compared and
here {\bf clumsy wording} we see comparisons between three surfers
and, in the case of wct08, many more.  It is natural to ask how much
information is contained in the observations where more than two
surfers are in the water.

We can directly calculate the determinant of the observed information
matrix for the likelihood function in
equation~\ref{complete_likelihood_function} at about $4.3\times
10^{71}$, as compared with~$2.9\times 10^{69}$ for the pairs
considered in isolation, a ratio of about $146.4$.  We are thus
motivated to consider the whole dataset, not just the part which
involves paired comparisons.  {\bf clumsy wording}.

@@

\section{Brazilian beach type prefereneces}

@@

One measure of the rising popularity of surfing is its inclusion in
the 2020 Summer Olympics, and it is natural to ask whether this would
affect the competition in some way.  One suggestion {\bf Tim?} would
be that competitors would become more focussed on beating compatriots
than surfers from other countries, the thinking being that admission
to a country's Olympic squad would be of more importance to a
competitor than their result in wct.  {\bf Tim this is clumsy
  wording}.  We would expect competitors to value their ranking when
restricted to within their country more than their ranking over all of
wct.  If this is the case then we would expect competitors to change
their behaviour if they are in the water with another surfer of the
same nationality.  Specifically, if we have three surfers in the water
of whom two are of one nationality and the third of another, we would
expect the compatriots to adopt a style calculated to beat their
countryman, possibly at the expense of performance relative to the odd
man out.  {\bf Tim this is poor wording we need to rephrase}.

Such hypotheses are difficult to test with conventional techniques as
competitors's strengths are nuisance parameters ~\cite{basu1977}.
However, the

>>>>

\title{Analysis of competitive surfing tournaments with generalized Bradley-Terry}

\section*{Abstract}
Here we apply a generalization of Bradley-Terry model to competitive
surfing, specifically the 2018 World Surf League.  We use
Bradley-Terry likelihoods to assess a number of hypotheses of interest
to the surfing community with formal statistical methods.  We go on to
study the ``Brazilian storm'' phenomenon using reified Bradley-Terry
likelihood functions in two ways: firstly, the supposed Brazilian
preference for point and beach break beaches; and secondly from the
perspective of tournament theory in which we test for competitors
modifying their strategy in the presence of compatriot rivals.

@@

Paired comparisons are particularly favourable to Bradley-Terry
because the likelihood function is exponential family and as such is
amenable to analysis via general linear models~\cite{turner2012}.
However, this is not applicable when more than two entities are
compared; and indeed here we see comparisons between three surfers
and, in the case of Jeep Surf Freshwater Pro (wct08), many more.  It
is natural to ask how much information is contained in the
observations where more than two surfers are in the water.

We can directly calculate the determinant of the observed information
matrix for the likelihood function in
equation~\ref{complete_likelihood_function} at about $4.3\times
10^{71}$, as compared with~$2.9\times 10^{69}$ for the pairs
considered in isolation, a ratio of about $146.4$.  This ratio shows
that there is considerable information loss in considering only the
paired comparisons.  We are thus motivated to consider the whole
dataset, not just the part which involves paired comparisons.

@@

\section{The Brazilian storm}

In surfing lore, the ``Brazilian storm'' refers to the emergence of
Brazil as a dominant nation of elite surfers.  The term was first used
to describe the fact six out of eight querterfinalists at 2011 Nike
Lowers Pro were Brazilian; it rose in public conciousness following
the victories of Medina and de Souza in the World Surf League
championships in 2014 and 2015 respectively.  The storm shows no signs
of abating, with commentators such as Douglas opining that this
generation ``represents a seismic shift for Brazilian competitors onto
an entirely new plane'' \cite{douglas2020}.

Here, we study two aspects of this sporting phenomenon: the suggested
Brazilian preference for beach break and point break reefs; and the
noncompetitive effect of compatriot rivalry.

\subsection{Brazilian beach type prefereneces}

@@

One measure of the rising popularity of surfing is its inclusion in
the 2020 Summer Olympics, and it is natural to ask whether this would
affect the competition in some way.  One suggestion {\bf Tim?} would
be that competitors would become more focussed on beating compatriots
than surfers from other countries, the thinking being that admission
to a country's Olympic squad would be of more importance to a
competitor than their result in wct.  {\bf Tim this is clumsy
  wording}.  We would expect competitors to value their ranking when
restricted to within their country more than their ranking over all of
wct.  If this is the case then we would expect competitors to change
their behaviour if they are in the water with another surfer of the
same nationality.  Specifically, if we have three surfers in the water
of whom two are of one nationality and the third of another, we would
expect the compatriots to adopt a style calculated to beat their
countryman, possibly at the expense of performance relative to the odd
man out.  {\bf Tim this is poor wording we need to rephrase}.  Such
hypotheses are difficult to test with conventional techniques as
competitors's strengths themselves are nuisance
parameters~\cite{basu1977}.
